---
title: "Introduction to R: Session 2"
date: "`r format.Date(Sys.Date(), format = '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    df_print: paged
---

```{r, include = FALSE}

knitr::opts_chunk$set(eval = TRUE,
                      message = FALSE,
                      warning = FALSE)

```

Original material developed by Abigail Page, Claire Dooley, Anushé Hassan, and Shammi Luhar.

**Updated (2022) by Alexis Robert and Sophie Meakin.**

# Introduction

Welcome to *Introduction to R* for MSc Epidemiology 2022-23!

## Contents

This session introduces the data visualisation with ggplot2.

## Notation

This exercise will contain examples of code and its output, as well as instructions to edit or write code yourself. We've used the following symbols to highlight certain instructions:

-   `r emoji::emoji("star")` A key piece of information e.g. a useful R function.
-   `r emoji::emoji("computer")` An instruction to copy-and-run code from the exercise, or write your own code, in your R Studio session.
-   `r emoji::emoji("question")` A question, related to the material or code output.

## Resources

* [The Epidemiologist R Handbook](https://www.epirhandbook.com/en/) - a comprehensive reference guide to using R for applied epidemiology and public health.
* [R for Data Science](https://r4ds.had.co.nz/index.html) - a general guide to using R for data science.
* [Stack Overflow](https://stackoverflow.com/questions/tagged/r) - general R questions answered by the community.

**For ggplot2**

* [Cheatsheet - ggplot2](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf) - a two-page summary of the key features and functions of ggplot2.
* [The R graph gallery](https://r-graph-gallery.com/) - a collection of charts made with R. Great for inspiration!
* [The Epidemiologist R Handbook: chapters 30-31](https://www.epirhandbook.com/en/ggplot-basics.html) - introduction to the basics ggplot2. Later chapters (32-39) cover also more complex topics.



# Recap: importing data

`r emoji::emoji("star")` We import a data file (e.g. .csv) using the `import()` function from the **rio** library, and tell R where to look for the file using the `here()` function from the **here** library.

The `here()` function works with your Rproj file to build a file path relative to the root folder; see Session 1 for more details.

`r emoji::emoji("computer")` Copy and paste the two chunks of code below into an R script and save it. This code loads the two libraries and imports the data `life_expectancy.csv` and assigns it to an object called `life_expectancy`:

```{r}
# Load rio and here libraries for importing data
library(rio)
library(here)
```

```{r}
# Import life_expectancy.csv
life_expectancy <- import(file = here("data",
                                      "life_expectancy.csv"))

```

# Data visualisation with ggplot2

We are going to use the ggplot2 library for data visualisation.

```{r}
# Load ggplot2 library
library(ggplot2)
```

## The basics of ggplot2

`r emoji::emoji("star")` There are three essential steps:

1.  Open the canvas with `ggplot()` and define which data set to use.
2.  Map variable (column) names to the aesthetics (features) of the plot, such as the x and y axes.
3.  Define the type of plot to use - e.g. a histogram, a scatter plot, a boxplot.

### Open the canvas

`r emoji::emoji("star")` We open the canvas with the `ggplot()` function, and define which data set to use with the `data` argument:

```{r}
ggplot(data = life_expectancy)
```

`r emoji::emoji("computer")` Copy and paste the code above into your R Studio console and run it.

An empty plot will open in the "Plots" tab of the bottom-right panel! The above command tells R we are starting a plot - but we will need to tell R *which variables* to plot (step 2) and *how* to plot them (step 3).

### Map data to plot aesthetics

Next, we need to tell R which variables (i.e. which columns) of our data we want to plot. To do this, we map column names from our data to plot aesthetics (features), such as the x and y axes.

`r emoji::emoji("star")` We map variables to aesthetics inside the `aes()` function, and assign this to the `mapping` argument of the `ggplot()` function.

In the following code, we map the column "health_expenditure_usd" to the x-axis, and "life_expectancy_years" to the y-axis:

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years))
```

`r emoji::emoji("computer")` Copy and paste the code above into your R Studio console and run it.

The variables will now appeared on their respective axes, and there are numbers indicating the range of the values. But there is no data yet because we still need to tell R *how* to plot the data (step 3).

### Define the plot type

Finally, we need to tell R *how* to plot our chosen data and variables.

`r emoji::emoji("star")` We define the plot type with a `geom_` function.

`r emoji::emoji("star")` We add the chosen `geom_` function by writing a plus symbol (`+`) after the end of the `ggplot()` function.

In the following code, we add the `geom_point()` function to make a scatter plot:

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years)) +
  geom_point()
```

`r emoji::emoji("computer")` Copy and paste the code above into your R Studio console and run it.

Congratulations, you've made your first plot in R! The `geom_point()` function adds points to the plot. Each point represents a row in the data `life_expectancy` where neither health expenditure nor life expectancy are missing. 

`r emoji::emoji("question")` There are many different "geoms" available, each for a different plot type. Look at some of the different plots on the [The R graph gallery](https://r-graph-gallery.com/). What are the names of some other "geom" functions?



## Example 1: scatter plot

In this section, we will walk through the steps to make the following plot. The plot shows of country-level life expectancy at birth in years versus health expenditure per capita in USD for 33 countries:

```{r, echo = FALSE, fig.width = 9, fig.height = 7}

ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years,
                     col = region)) +
  geom_point(size = 4, shape = "circle") +
  scale_y_continuous(limits = c(73, 85),
                     breaks = seq(from = 73, to = 85, by = 4)) +
  scale_color_brewer(palette = "Set2") +
  labs(x = "Health expenditure (USD per capita)",
       y = "Life expectancy (years)",
       col = "Region",
       title = "Life expectancy at birth vs. health expenditure, 2015") +
  theme_minimal() +
  theme(legend.position = "top")

```

The code to make this plot will begin with the code we saw in the previous section.

`r emoji::emoji("computer")` Copy and paste the following code into your R script.

```{r, eval = FALSE, fig.width = 9, fig.height = 7}

ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years)) +
  geom_point()

```


### Change the size, shape and colour of points

First, we look at how to change the size, shape and colour of the points in a scatter plot.

#### Fixed size, shape and colour {-}

`r emoji::emoji("star")` We can change the fixed attribute of the points by setting the arguments `size`, `shape` and `colour` inside the geom function.

In the following code we set the point size to 10, the shape to a diamond and the colour to "hotpink":

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years)) +
  geom_point(size = 10, shape = "diamond", color = "hotpink")
```

* The `size` argument changes the size of the points. The argument takes positive numeric values, where bigger values equate to bigger points.

* The `shape` argument changes the shape of the points. This [ggplot2 vignette](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#sec:shape-spec) has more details about the values that this argument can take.

`r emoji::emoji("question")` Read the ggplot2 vignette section about `shape`. What other values can this take?

* The `color` argument changes the colour of the points. The argument takes values from the output of the `colors()` function.

`r emoji::emoji("question")` Run the `colours()` function in your R Studio console. What other values can this take?

`r emoji::emoji("computer")` Edit the code in your R script to change the default size, shape and colour of the points.

#### Dynamic colours {-}

`r emoji::emoji("star")` We can change the colour of the points according to values in the data by mapping a column name to the `color` aesthetic inside the `aes()` function.

In the following code, we map the column "region" to the `color` argument:

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years,
                     color = region)) +
  geom_point()
```


### Other optional plot changes

In this section we look at how to do the following:

-   Change the limits and breaks on a numeric axis
-   Change the colour palette
-   Add axis labels and a title
-   Change the plot theme (general appearance)
-   Move the legend

`r emoji::emoji("computer")` Copy and paste the code below into a script in your R Studio session.

```{r, eval = FALSE}

ggplot(data = life_expectancy,
       mapping = aes(x = health_expenditure_usd,
                     y = life_expectancy_years,
                     col = region)) +
  geom_point(size = 4, shape = "diamond") +
  
  # Update the y-axis limits and breaks
  scale_y_continuous(limits = c(60, 90),
                     breaks = seq(from = 60, to = 90, by = 10)) +
  
  # Update the colour palette
  scale_color_brewer(palette = "Set2") +
  
  # Add axis labels, legend title (color) and plot title
  labs(x = "health_expenditure_usd",
       y = "life_expectancy_years",
       color = "Region",
       title = "My first ggplot!") +
  
  # Update the theme
  theme_minimal() +
  
  # Move the legend
  theme(legend.position = "right")

```

#### Change the limits and breaks on a numeric axis {-}

`r emoji::emoji("star")` Use the `scale_y_continuous()` function to change a numeric y axis.

We can update the limits using the `limits` argument, and we can update the breaks using the `breaks` argument.

In the code above, the y-axis lower and upper limits are 60 and 90, respectively (`limits = c(60, 90)`), and the axis breaks start at 60, end at 90 and increase by 10 (`breaks = seq(from = 60, to = 90, by = 10)`).

`r emoji::emoji("computer")` Edit your code to make the following changes to the y-axis limits and breaks: change the y-axis lower limit to 73 and the upper limit to 85, and change the y-axis breaks to start at 73, end at 85 and increase by 4.

#### Change the colour palette {-}

Use the `scale_color_brewer()` function to change the colour palette used for a dynamic colour aesthetic.

We can choose an appropriate palette by defining the `palette` argument.

`r emoji::emoji("question")` Read the documentation for `scale_color_brewer()`. What are the names of some of the other qualitative palettes available?

#### Change axis labels and add a plot title {-}

`r emoji::emoji("star")` Use the `labs()` function to define axis labels, the plot title, and more.

The arguments `x` and `y` define the x- and y-axis labels, respectively, while the `title` argument defines the plot title.

In the code above, the x-axis title is `"health_expenditure_usd"`, the y-axis title is `"life_expectancy_years"`, and the title of the plot is `"My first ggplot!"`.

`r emoji::emoji("computer")` Edit your code to give the x-axis, y-axis and plot title more informative values. If you're stuck on *what* to write, you can refer to the plot at the start of the section if you want to, or you can choose your text values.

`r emoji::emoji("question")` Run `?labs()` in your R Studio console and read the documentation. What other text can you add or edit with the `labs()` function?

#### Change the plot theme {-}

`r emoji::emoji("star")` One of the easiest ways to make changes to the general design of a plot is with a complete theme, using a function of the format `theme_()`.

A complete theme will adjust many of the features of the plot, such as the background colour, the axes, whether there are gridlines or not, and text size, colour and font. One example of a complete theme is the function `theme_minimal()`.

`r emoji::emoji("computer")` Read the ggplot2 documentation page about [complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html). Edit your code to try some other complete themes.

#### Change the legend position {-}

`r emoji::emoji("star")` Individual elements of the plot appearance can be adjusted within the `theme()` function. For example, we can move the legend position with `legend.position` argument.

In the code above, the legend is on the `"right"`.

`r emoji::emoji("computer")` Edit your code to move the legend to the `"top"` or `"bottom"` of the plot.



## Example 2: histogram

We are going to walk you through the steps to make this histogram of life expectancy at birth, using the data in `life_expectancy`:

```{r, echo = FALSE, fig.width = 9, fig.height = 7}

ggplot(data = life_expectancy,
         mapping = aes(x = life_expectancy_years,
                       fill = region)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ region) +
  scale_y_continuous(breaks = seq(from = 0, to = 16, by = 4)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Life expectancy (years)",
       y = "Number of countries",
       fill = "Region",
       title = "Life expectancy at birth, 2015") +
  theme_minimal() +
  theme(legend.position = "none")

```

### Change the bin width

```{r}

ggplot(data = life_expectancy,
       mapping = aes(x = life_expectancy_years)) +
  geom_histogram(binwidth = 1)

```

### Change the color and fill of bars

#### Static color and fill {-}

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = life_expectancy_years)) +
  geom_histogram(binwidth = 1, col = "seagreen", fill = "tomato")
```

### Dynamic fill {-}

```{r}
ggplot(data = life_expectancy,
       mapping = aes(x = life_expectancy_years,
                     fill = region)) +
  geom_histogram(binwidth = 1)
```


### Other optional plot changes

We are going to cover how to:

-   Break a plot into subplots (facets)
-   Change the fill colour palette
-   Remove the legend

You will also see the same functions used earlier to change the y-axis scale, to add axis labels and a plot title, and to change the theme.

`r emoji::emoji("computer")` Copy and paste the code below into a script in your R Studio session.

```{r, eval = FALSE}

ggplot(data = life_expectancy,
         mapping = aes(x = life_expectancy_years,
                       fill = region)) +
  geom_histogram(binwidth = 1) +
  
  # Divide the plot into subplots (facets)
  facet_wrap(~ region) +
  
  # Update the y-axis limits and breaks
  scale_y_continuous(breaks = seq(from = 0, to = 16, by = 4)) +
  
  # Update the fill colour palette
  scale_fill_brewer(palette = "Accent") +
  
  # Add axis labels, legend title (fill), and plot title
  labs(x = "Life expectancy (years)",
       y = "Number of countries",
       fill = "Region",
       title = "Life expectancy at birth, 2015") +
  
  # Update the theme
  theme_minimal() +
  
  # Remove the legend
  theme(legend.position = "none")

```

## Other plot types

# Extras

### 2.11 Problem shooting

It is really common to run a line of code and it does not work. Normally it is a misplaced character, or an unclosed bracket. Check that every ( has a ) and every " has a ".

Capitals also matter in R - note that Sepal.Length will work but sepal.length will not!

Sometimes you will run the code and nothing happens. Check the left-hand of your console: if its a +, it means that R does not think you have typed a complete expression and its waiting for you to finish it. In this case, its usually easy to start from scratch again by pressing ESCAPE to abort processing the current command.

One common problem when creating ggplot2 graphics is to put the + in the wrong place: it has to come at the end of the line, not the start.

If you are still stuck, try the help. You can get help about any R function by running ?function_name in the console, or selecting the function name and pressing F1 in RStudio. If the help does not seem that helpful, skip down to the examples and look for code that matches what you are trying to do.

If that does not help, carefully read the error message. Sometimes the answer will be buried there! But when you are new to R, the answer might be in the error message but you do not yet know how to understand it. Another great tool is Google: try googling the error message, as it is likely someone else has had the same problem, and has gotten help online.

### 2.12 Resources

<https://www.rstudio.com/resources/cheatsheets/> - cheatsheet for ggplot2

<https://r4ds.had.co.nz/data-visualisation.html> - R for Data Science

<http://www.cookbook-r.com/Graphs/> - Cookbook for R Graphics
